abxD01 Paper

### 2015-03-10 PAPER.Rmd
Add to discussion relate to humans

Throughout our analysis, we observed examples of the significance of the microbiome as a whole over any individual bacterial species in contributing to resistance. Correlation analysis identified whole families with significant negative or positive relationships with *C. difficile*. However, correlation analysis alone was insuficient in identifying all informative bacteria. Random forest modeling allowed us to identify several bacterial OTUs that did not have significant correlations with *C. difficile* yet improved model accuracy. *Turicibacter* and *Akkermansia*, for example, both had non-significant correlations with *C .difficile* yet were more important for model accuracy over OTUs with stronger correlations (*Figure 5BC*). While these bacteria may not have important direct effects on *C. difficile*, they may interact positively or negatively with bacteria that do have strong relationships with *C. difficile*. Through the model analysis, we observed that no single bacteria was able to consistently explain the subsequent level of *C. difficile* observed. For example, the *Porphyromonadaceae* OTU3, along with the *Lachnospiraceae* OTU39, were the two most important OTUs in the model with strong negative correlations with *C. difficile*. The majority of microbiotas which had over 1% relative abundance of OTU3 had no detectable *C. difficile* (*Figure 5C*). However, a small group of microbiotas with greater than 1% OTU3 abundance, which included several streptomycin (0.1 mg/ml) treated, ampicillin treated (both 1 and 6 days recovery), and cefoperazone treated (both 0.1 mg/ml and 0.3 mg/ml) microbiotas, were not fully protected by the presence of this OTU alone and ranged in *C. difficile* levels from 10<sup>2</sup> to 10<sup>6</sup> CFU/g feces. Similarly, having low amounts of one "protective" bacteria such as *Lachnospriaceae* OTU39 does not automatically render the host susceptible to colonization. For instance streptomycin treated (0.1 mg/ml) mice did not harbor *Lachnospiraceae* OTU39, yet had undetectable levels of *C. difficile*. Thus the loss or absence of one protective bacterial species can be compensated by others, either related species or others with functional redundancy (as in redundancy in their protective abilities).

Furthermore, the dynamics of *C. difficile* colonization does not appear to be dependent soley on the loss of protective species. Both *Escherichia* OTU11 and *Streptococcus* OTU78 were positively correlated with *C. difficile* colonization and important in the random forest model. Notably, *Escherichia* OTU11 was among the strongest positively correlated OTUs with *C. difficile* across all antibiotic treatments, except for streptomycin. In fact, while streptomycin treatment (5 mg/ml) lead to high *C. difficile* colonization, these communities had very little, if any, *Escherichia* OTU11. This demostrates that while *Escherichia* may aid in colonization of *C. difficile*, it is not essential. These observations lead to a model of colonization by *C. difficile* in which antibiotic perturbation leads to loss of protective species. This in turn eliminates competition, frees resources, and eliminates direct or indirect inhibitory effects of bacterial products (e.g. bacteriocins or short chain fatty acids) made by these protective bacteria. While this may be suitable enough for *C. difficile* to colonize, other opportunistic species such as *Escherichia coli* and other *Proteobacteria* may also take advantage of these microbiota perturbations. Increased colonization by these bacteria can increase inflammation, produce a new profile of metabolites and bacterial products, reduce the mucus layer,  and deoxygenate the microoxic tiers nearing the epithelium. These changes in the gut milieu may be advantageous for *C. difficile* and other gut pathogens. Inflammation, for instance, has been shown to be advantageous for *Salmonella tyhpumurium* to further disrupt the indigenous microbiota []. Inflammation also leads to a shift in the availability of resources, such as nitrites, which can be used as an alternative resource by *Salmonella typhimurium* []. *E.coli* is capable of using resources found in the mucus layer while simultaneously deoxygenating this region []. This reduces the spatial protection of the mucus layer, allowing bacteria, including strictly anaerobic bacteria, and bacterial products (e.g. toxins) to more readily reach the epithelial layer.

>>>>

Correlation analysis identified whole families with significant negative or positive relationships with *C. difficile*.  Many of the correlaions we observed were not surprising given other studies looking at the relationship between *C. difficile* and the microbiome. For instance, monocolonization of germfree mice with either a *Lachnospiraceae* species or *E. coli* were tested for their resistance to *C. difficile* colonization [Reeves]. Lachnospiraceae was able to decrease the levels of *C. difficile* observed, while *E. coli* had similar colonization levels compared to germfree mice given *C. difficile* alone. Association studies in humans with and without CDI in hospitals have also highlighted several bacteria found in this study, including *Lachnospiraceae*, *Rumicoccaceae*, *Clostridia*, and *Alistipes* as decreased and *E. coli* and *Streptococcus* as increased in subjects with CDI [Schubert, Antharam]. While we observe overlaps in bacteria with potentially protective roles in both mice and humans at broad taxonomy levels, there are notable differences at refined levels. *Porphyromonadaceae*, for example, are more common among mice than humans, where instead there are strong negative associations of *C. difficile* with *Bacteroides* [Schubert]. However, correlation analysis alone was insuficient in identifying all informative bacteria. Random forest modeling allowed us to identify several bacterial OTUs that did not have significant correlations with *C. difficile* yet improved model accuracy. *Turicibacter* and *Akkermansia*, for example, both had non-significant correlations with *C .difficile* yet were more important for model accuracy over OTUs with stronger correlations (*Figure 5BC*). While these bacteria may not have important direct effects on *C. difficile*, they may interact positively or negatively with bacteria that do have strong relationships with *C. difficile*. Few studies attempt to model the microbiome to predict disease or estimate health [Pamer, clostridium paper, ].

Throughout our analysis, we observed examples of the significance of the microbiome as a whole over any individual bacterial species in contributing to resistance. No single bacteria was able to consistently explain the subsequent level of *C. difficile* observed. For example, the *Porphyromonadaceae* OTU3, along with the *Lachnospiraceae* OTU39, were the two most important OTUs in the model with strong negative correlations with *C. difficile*. The majority of microbiotas which had over 1% relative abundance of OTU3 had no detectable *C. difficile* (*Figure 5C*). However, a small group of microbiotas with greater than 1% OTU3 abundance, which included several streptomycin (0.1 mg/ml) treated, ampicillin treated (both 1 and 6 days recovery), and cefoperazone treated (both 0.1 mg/ml and 0.3 mg/ml) microbiotas, were not fully protected by the presence of this OTU alone and ranged in *C. difficile* levels from 10<sup>2</sup> to 10<sup>6</sup> CFU/g feces. Similarly, having low amounts of one "protective" bacteria such as *Lachnospriaceae* OTU39 does not automatically render the host susceptible to colonization. For instance streptomycin treated (0.1 mg/ml) mice did not harbor *Lachnospiraceae* OTU39, yet had undetectable levels of *C. difficile*. Thus the loss or absence of one protective bacterial species can be compensated by others, either related species or others with functional redundancy (as in redundancy in their protective abilities).

Furthermore, the dynamics of *C. difficile* colonization does not appear to be dependent soley on the loss of protective species. Both *Escherichia* OTU11 and *Streptococcus* OTU78 were positively correlated with *C. difficile* colonization and important in the random forest model. Notably, *Escherichia* OTU11 was among the strongest positively correlated OTUs with *C. difficile* across all antibiotic treatments, except for streptomycin. In fact, while streptomycin treatment (5 mg/ml) lead to high *C. difficile* colonization, these communities had very little, if any, *Escherichia* OTU11. This demostrates that while *Escherichia* may aid in colonization of *C. difficile*, it is not essential. These observations lead to a model of colonization by *C. difficile* in which antibiotic perturbation leads to loss of protective species. This in turn eliminates competition, frees resources, and eliminates direct or indirect inhibitory effects of bacterial products (e.g. bacteriocins or short chain fatty acids) made by these protective bacteria. [can talk about change in gut with abx--loss of some, gain of others that are resistant to abx or grow up quickly after, and direct effects of abx... last weeks jclub paper; sonnenburg has 2 papers with nutrients/metbolites and cdiff related to the microbiome]

While this may be suitable enough for *C. difficile* to colonize, other opportunistic species such as *Escherichia coli* and other *Proteobacteria* may also take advantage of these microbiota perturbations. Increased colonization by these bacteria can increase inflammation, produce a new profile of metabolites and bacterial products, reduce the mucus layer,  and deoxygenate the microoxic tiers nearing the epithelium. These changes in the gut milieu may be advantageous for *C. difficile* and other gut pathogens. Inflammation, for instance, has been shown to be advantageous for *Salmonella tyhpumurium* to further disrupt the indigenous microbiota []. Inflammation also leads to a shift in the availability of resources, such as nitrites, which can be used as an alternative resource by *Salmonella typhimurium* []. *E.coli* is capable of using resources found in the mucus layer while simultaneously deoxygenating this region []. This reduces the spatial protection of the mucus layer, allowing bacteria, including strictly anaerobic bacteria, and bacterial products (e.g. toxins) to more readily reach the epithelial layer.
